Mainstreaming Scientific Racism

### 2022-10-19 talk/talk.qmd
incorporate UMAP vis into talk

# Race science

## Some definitions

@@

- This talk will focus on narrow-sense race science

## Race science in the 20th century: Pre-*Brown* {.smaller}

:::: columns 
::: {.column width="30%"}
![[Charles Davenport (1866-1944)](https://en.wikipedia.org/wiki/Charles_Davenport), circa 1929. Source: <https://en.wikipedia.org/wiki/File:Charles_Benedict_Davenport.jpg>](img/Charles_Benedict_Davenport.jpg){width="100%"}
:::

::: {.column width="30%"}
![[Franz Boas (1858-1942)](https://en.wikipedia.org/wiki/Franz_Boas), circa 1915. Source: <https://en.wikipedia.org/wiki/File:FranzBoas.jpg>](img/FranzBoas.jpg)
:::

::: {.column width="30%"}
![[Margaret Mead (1901-1978)](https://en.wikipedia.org/wiki/Margaret_Mead), 1948. Source: <https://en.wikipedia.org/wiki/File:Margaret_Mead_(1901-1978).jpg>](img/Margaret_Mead_(1901-1978).jpg)
:::
::::

@@

:::: columns
::: {.column width="30%"}
![[Reginald Ruggles Gates (1882-1962)](https://en.wikipedia.org/wiki/Reginald_Ruggles_Gates), 1921. Source: <https://en.wikipedia.org/wiki/File:Reginald_Ruggles_Gates_Portrait.jpg>](img/Reginald_Ruggles_Gates_Portrait.jpg)
:::

::: {.column width="30%"}
![[Henry Garrett (1894-1973)](https://en.wikipedia.org/wiki/Henry_Garrett_(psychologist)). Source: <https://www.psychometricsociety.org/post/past-present-and-incoming-presidents>, citing American Psychologist (1946), 1(9), 371](img/henry_e_garrett.png)
:::

::: {.column width="30%"}
![[George Robert Gayre of Gayre and Nigg (1907-1996)](https://en.wikipedia.org/wiki/Robert_Gayre). Source: <https://www.patrickcomerford.com/2020/07/robert-gair-racist-from-rathmines-who.html>, original uncredited](img/Robert Gayre profile.jpg)
:::
::::

@@

> [E]veryone agrees that races or ethnic groups exist and that they are in fact the raw material by means of which human evolution has taken place. (Gates, *MQ* 1:1 [1960], 11)

# Race science as agnotology and echo chamber

@@

@@

- 34k journal articles
- Published 1960-2010
- 3 journals published by American Psychological Association not available
- PID is missing many relevant book reviews and essays
	- eg, [Rushton's review of Lynn's book on Pioneer](https://www.sciencedirect.com/science/article/pii/S0191886901001337) *[confirm not included]*

@@

## Gould's critique of factor analysis

@@

These words tend to be used together; **but why?**\
These test items tend to be answered in the same way; **but why?**

>>>>

# Race science in the 20th century

<!--
## Some definitions

@@

- This talk will focus on narrow-sense race science
-->

## Before *Brown* {.smaller}

:::: columns 
::: {.column width="30%"}
![[Charles Davenport (1866-1944)](https://en.wikipedia.org/wiki/Charles_Davenport), circa 1929. [Source](https://en.wikipedia.org/wiki/File:Charles_Benedict_Davenport.jpg)](img/Charles_Benedict_Davenport.jpg){width="100%"}
:::

::: {.column width="30%"}
![[Franz Boas (1858-1942)](https://en.wikipedia.org/wiki/Franz_Boas), circa 1915. [Source](https://en.wikipedia.org/wiki/File:FranzBoas.jpg)](img/FranzBoas.jpg)
:::

::: {.column width="30%"}
![[Margaret Mead (1901-1978)](https://en.wikipedia.org/wiki/Margaret_Mead), 1948. [Source](https://en.wikipedia.org/wiki/File:Margaret_Mead_(1901-1978).jpg)](img/Margaret_Mead_(1901-1978).jpg)
:::
::::

@@

:::: columns
::: {.column width="30%"}
![[Reginald Ruggles Gates (1882-1962)](https://en.wikipedia.org/wiki/Reginald_Ruggles_Gates), 1921. [Source](https://en.wikipedia.org/wiki/File:Reginald_Ruggles_Gates_Portrait.jpg)](img/Reginald_Ruggles_Gates_Portrait.jpg)
:::

::: {.column width="30%"}
![[Henry Garrett (1894-1973)](https://en.wikipedia.org/wiki/Henry_Garrett_(psychologist)). [Source](https://www.psychometricsociety.org/post/past-present-and-incoming-presidents), citing American Psychologist (1946), 1(9), 371](img/henry_e_garrett.png){width="100%"}
:::

::: {.column width="30%"}
![[George Robert Gayre of Gayre and Nigg (1907-1996)](https://en.wikipedia.org/wiki/Robert_Gayre). [Source](https://www.patrickcomerford.com/2020/07/robert-gair-racist-from-rathmines-who.html), original uncredited](img/Robert Gayre profile.jpg)
:::
::::

@@

## *MQ* was founded as race science

> [[E]veryone agrees that races or ethnic groups exist and that they are in fact the raw material by means of which human evolution has taken place]{.red} ....

> A scientific approach to racial questions, [free from sentiment, propaganda or bias]{.blue}, may lead to a more reasonable attitude to the problems which agitate considerable sections of mankind. This need for a calmer approach to problems of race on the part of [civilised peoples as well as more primitive tribes]{.red} makes it desirable that a journal should exist in which the problems of racial origin and racial relationships can be quietly discussed, [without rancour or bigotry and with the primary aim of elucidating facts]{.blue}. (Gates, *MQ* 1:1 [1960], 11)

# Race science agntology

@@

## Did PF and MQ function as a race science echo chamber?

- Empirical, text-mining approach
- Identify Pioneer-funded researchers
- Identify journals that published their work
- Analyze all articles published in those journals, 1960-2010

@@

- 34k journal articles
- Published 1960-2010

\
- 3 journals published by American Psychological Association not available
- PID is missing many relevant book reviews and essays
	- eg, [Rushton's review of Lynn's book on Pioneer](https://www.sciencedirect.com/science/article/pii/S0191886901001337) *[confirm not included]*

@@

<!-- 
## Gould's critique of factor analysis

@@

These words tend to be used together; **but why?**\
These test items tend to be answered in the same way; **but why?**
 -->

# Mainstreaming race science

## Umap interactive

```{r}
#| out-width: 1440px
#| out-height: 1080px
#| fig-cap: "UMAP visualization of Hellinger similarity, medium vocabulary, k = 30"
library(readr)
library(here)
library(plotly)

vis_dir = here('out')

umap_vis = read_rds(here(vis_dir, '07_md_30.Rds'))
umap_vis
```
